Concurrent Programming with CXL

### 2020-06-16 book/paper.tex
More index terms

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes concurrently read and update shared variables
\index{shared variable}
and synchronize with one another makes programs more complicated than
programs where only one thing happens at a time.  Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential program is mostly \emph{deterministic}.
\index{determinism}
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
The output of running concurrent programs depends on how the various
processes are interleaved.  Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\index{Heisenbug}
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically}
\index{atomicity}
because there is no other activity
interfering with their execution.  Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other instructions
may update memory locations as a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

@@

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved in unspecified and often random ways.
A program is correct if it works for any interleaving.
In fact, CXL will try all possible interleavings of the processes
executing machine instructions.

@@

The state of a process is called a \emph{context} (aka continuation):
\index{context}
\index{continuation}
it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag}) of
contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty dictionary, \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
CXL values are unbounded trees.
Both the memory and the register of a process always contain
a dictionary that maps atoms to CXL values.  We call this a \emph{directory}.
\index{directory}
A directory represents the state of a collection of variables named by the atoms.
%
Because directories are CXL values themselves,
directories can be organized into a tree.
Each node in a directory tree is then identified
by a sequence of atoms, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address} of a CXL value, and it is relative to the
``root'' of the directory tree.

@@

\chapter{Critical Sections}
\index{critical section}
\index{critical region}

@@

Hopefully you have started thinking of how to solve the concurrency
problem and you may already have prototyped some solutions.
In this chapter we will go through a few reasonable but broken attempts.
At the heart of the problem is that we would like make sure that, when
the \texttt{count} variable is being updated, no other process is
trying to do the same thing.  This is called a \emph{critical section}
(aka critical region)~\cite{EWD123}: a
set of instructions where only one process is allowed to execute at a
time.

@@

The CXL machine starts in an initial state in which there is only
one process and its program counter is~0.  A \emph{trace} is a
sequence of steps starting from the initial state.
When making a step, there are two sources of non-determinism
\index{non-determinism}
in CXL.
One is when
there is more than one process that can make a step.  The other is
when a process executes a \texttt{choose} operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable} if it is either the initial state
or it can be reached from the initial state through a trace.
A state is final
when there are no processes left to make state changes.

A \emph{property} describes a subset of
states.  In other words, a property is equivalent to a subset of the states.
\emph{Initial}, \emph{final}, and \emph{reachable}, and \emph{unreachable}
are all examples of properties.
Figure~\ref{fig:states} depicts a Venn diagram of various properties
and a trace.
An \emph{invariant} is a property that holds over all reachable states.
We want to show that mutual exclusion is an invariant.
In other words, we want to show that the set of reachable states is a subset
of the set of states where there is at most one process in the critical
section.

@@

One candidate for such an invariant is mutual exclusion itself.
After all, it certainly holds over the initial state.
And as CXL has already determined, mutual exclusion is an invariant:
it holds over every \emph{reachable state}.
Unfortunately, it is not an \emph{inductive} invariant.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 at the start of the \texttt{while} loop.
Also imagine that in this particular state $\mathtt{turn} = 1$.  Now let
process process~1 make a sequence of steps.  Because $\mathtt{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, violating mutual exclusion.
So mutual exclusion is an invariant, but not an inductive invariant.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

CXL is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In CXL, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this (as well as some other uses), CXL supports \emph{pointers}
\index{pointer}
to shared variables.
If \texttt{x} is a shared variable, then the expression \texttt{\&(x)}
(the parentheses are mandatory) is a pointer to \texttt{x}
(also known as the \emph{address} of \texttt{x}).
Conversely, if \texttt{p} is a pointer to a shared variable, then the
expression \texttt{\^{}p} is the value of the shared variable.

@@

Figure~\ref{fig:uplock} showed a faulty attempt at solving mutual
exclusion using a lock.  The problem with the implementation of the
lock is that checking the lock and setting it if it is available is
not \emph{atomic}.  Thus multiple processes contending for the lock
can all ``grab the lock'' at the same time.  While Peterson's
algorithm gets around the problem, it is not efficient, especially
if generalized to multiple processes.  Instead, multi-core processors provide
so-called \emph{interlock instructions}: special machine instructions
that can read memory and then write it in an indivisible step.

@@

We call a process \emph{blocked}
\index{blocked}
in a state if the process is in
an \emph{infinite loop} in which the process does not update the
shared memory.
A process trying to
acquire a test-and-set spinlock held by another process is a good example
of a process being blocked: the process only reads the shared memory
but does not change it.
The only way to break out of the infinite loop is if another process
changes the shared memory.
The \texttt{while} loop in Peterson's algorithm, in case its condition
does not hold, is another example of a process being blocked.

@@

\texttt{unlock} checks to see if any processes are waiting to get the lock.
If so, it uses the \texttt{head} and \texttt{tail}
methods from the \texttt{list} module to resume the first process that got
suspend and to remove its context from the list.
Selecting the first process is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
\texttt{choose} to pick an arbitrary one.  Selecting the first is a common
choice in lock implementations as it prevents \emph{starvation}: every process
gets a chance to obtain the lock (assuming every process eventually releases
it).  Chapter~\ref{ch:starvation} will talk more about starvation.

@@

\chapter{Reader/Writer Locks using Busy Waiting}
\index{reader/writer lock}
\index{busy waiting}

@@

While correct, it is not considered a good solution.
The solution is an example of what is called \emph{busy-waiting}
(aka spin-waiting):
processes spin in a loop until some desirable application-level condition is met.
The astute reader might wonder if obtaining the mutex itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the mutex is available
(see Figure~\ref{fig:spinlocks}).

@@

A good example of a resource with multiple units is the
so-called \emph{bounded buffer} (aka \emph{ring buffer}).
It is essentially
a queue implemented using a circular buffer of a certain length and two pointers:
one where items are inserted and one from where items are extracted.  If the
buffer is full, processes that want to add more items have to wait.  In the usual
formulation, processes that want to extract an item when the buffer is empty
also have to wait.
This problem is known as the ``Producer/Consumer Problem.''

@@

Note the quite different usage of the \texttt{l\_} semaphores and \texttt{n\_}
semaphores.
People often call semaphores that are used as locks \emph{binary semaphores},
as their counters only take on the values 0 and 1.
They protect critical sections.
The second type of semaphores are called \emph{counting semaphores}.
They can be used to send signals between processes.
However, binary and counting semaphores are implemented the same way.

@@

In this chapter, we will present a version of a reader/writer lock
implementation that solves this type of starvation, but won't eliminate
the type of starvation.
In particular, the probability that either a reader or writer gets starved
will be exponentially diminishing.
We shall call such a solution \emph{fair} (although it does not quite
match the usual formal nor vernacular concepts of fairness).

@@

The actor synchronization model is popular in a variety of programming
languages, including Erlang and Scala.  Actor support is also available
through popular libraries such as Akka, which is available for various
programming languages.  In Python, Java, and C/C++,
actors can be easily emulated using threads and \emph{synchronized queues}
(aka \emph{blocking queues}) for messaging.
Each thread would have one such queue for receiving messages.
Dequeuing from an empty synchronized queue blocks the thread until
another thread enqueues a message on the queue.

@@

The solution is not practical because the \texttt{items} array must be
unbounded if an unbounded number of items may be produced.  Each slot
in the array is only used once, which is inefficient.  Also, the \texttt{inc}
and \texttt{swap} interlock instructions are not universally available
on existing hardware.
However, in the literature there are many examples of practical
non-blocking (aka \emph{wait-free}) synchronization algorithms~\cite{}.

@@

An \emph{lvalue} (short for left hand value of an assignment statement)
is something that can be assigned.  This can be a
shared variable, a process variable, or a dereferenced pointer
variable.  It can also be indexed.  Examples include
\texttt{var}, \texttt{var[e]}, and \texttt{\^{}p}.

@@

A context is considered \emph{blocked} if there does not exist a
sequence instructions that the process can execute that either writes
to a shared variable (even if that leaves the shared variable unchanged)
or terminates the process.

>>>>

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes concurrently read and update shared variables
\index{shared variable}
and synchronize with one another makes programs more complicated than
programs where only one thing happens at a time.  Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential
\index{sequential}
program is mostly \emph{deterministic}.
\index{determinism}
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
The output of running concurrent programs depends on how the various
processes are interleaved.  Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\index{Heisenbug}
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically}
\index{atomicity}
because there is no other activity
interfering with their execution.  Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other instructions
may update memory locations as a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

@@

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved
\index{interleaving}
in unspecified and often random ways.
A program is correct if it works for any interleaving.
In fact, CXL will try all possible interleavings of the processes
executing machine instructions.

@@

The state of a process is called a \emph{context} (aka continuation):
\index{context}
\index{continuation}
it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag})
\index{bag}
\index{multiset}
of contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty dictionary, \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
CXL values are unbounded trees.
Both the memory and the register of a process always contain
a dictionary that maps atoms to CXL values.  We call this a \emph{directory}.
\index{directory}
A directory represents the state of a collection of variables named by the atoms.
%
Because directories are CXL values themselves,
directories can be organized into a tree.
Each node in a directory tree is then identified
by a sequence of atoms, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address}
\index{address}
of a CXL value, and it is relative to the
``root'' of the directory tree.

@@

\chapter{Critical Sections}

@@

Hopefully you have started thinking of how to solve the concurrency
problem and you may already have prototyped some solutions.
In this chapter we will go through a few reasonable but broken attempts.
At the heart of the problem is that we would like make sure that, when
the \texttt{count} variable is being updated, no other process is
trying to do the same thing.  This is called a \emph{critical section}
(aka critical region)~\cite{EWD123}:
\index{critical section}
\index{critical region}
a set of instructions where only one process is allowed to execute at a
time.

@@

The CXL machine starts in an initial state in which there is only
one process and its program counter is~0.  A \emph{trace}
\index{trace}
is a sequence of steps starting from the initial state.
When making a step, there are two sources of non-determinism
\index{non-determinism}
in CXL.
One is when
there is more than one process that can make a step.  The other is
when a process executes a \texttt{choose} operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable}
\index{reachable state}
if it is either the initial state
or it can be reached from the initial state through a trace.
A state is final
when there are no processes left to make state changes.

A \emph{property}
\index{property}
describes a subset of
states.  In other words, a property is equivalent to a subset of the states.
\emph{Initial}, \emph{final}, and \emph{reachable}, and \emph{unreachable}
are all examples of properties.
Figure~\ref{fig:states} depicts a Venn diagram of various properties
and a trace.
An \emph{invariant}
\index{invariant}
is a property that holds over all reachable states.
We want to show that mutual exclusion is an invariant.
In other words, we want to show that the set of reachable states is a subset
of the set of states where there is at most one process in the critical
section.

@@

One candidate for such an invariant is mutual exclusion itself.
After all, it certainly holds over the initial state.
And as CXL has already determined, mutual exclusion is an invariant:
it holds over every reachable state.
Unfortunately, it is not an \emph{inductive} invariant.
To see why, we need to consider an \emph{unreachable} state.
It is easy to construct one: let process~0 be at label \texttt{@cs}
and process~1 at the start of the \texttt{while} loop.
Also imagine that in this particular state $\mathtt{turn} = 1$.  Now let
process process~1 make a sequence of steps.  Because $\mathtt{turn} = 1$,
process~1 will break out of the while loop and also enter the critical
section, violating mutual exclusion.
So mutual exclusion is an invariant, but not an inductive invariant.
Doing a inductive proof with an invariant that is not inductive is usually
much harder than doing one with an invariant that is.

@@

CXL is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In CXL, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this (as well as some other uses), CXL supports \emph{pointers}
\index{pointer}
to shared variables.
If \texttt{x} is a shared variable, then the expression \texttt{\&(x)}
(the parentheses are mandatory) is a pointer to \texttt{x}
(also known as the \emph{address}
\index{address}
of \texttt{x}).
Conversely, if \texttt{p} is a pointer to a shared variable, then the
expression \texttt{\^{}p} is the value of the shared variable.

@@

Figure~\ref{fig:uplock} showed a faulty attempt at solving mutual
exclusion using a lock.  The problem with the implementation of the
lock is that checking the lock and setting it if it is available is
not \emph{atomic}.  Thus multiple processes contending for the lock
can all ``grab the lock'' at the same time.  While Peterson's
algorithm gets around the problem, it is not efficient, especially
if generalized to multiple processes.  Instead, multi-core processors provide
so-called \emph{interlock instructions}:
\index{interlock instruction}
special machine instructions
that can read memory and then write it in an indivisible step.

@@

We call a process \emph{blocked}
\index{blocked process}
in a state if the process is in
an \emph{infinite loop} in which the process does not update the
shared memory.
A process trying to
acquire a test-and-set spinlock held by another process is a good example
of a process being blocked: the process only reads the shared memory
but does not change it.
The only way to break out of the infinite loop is if another process
changes the shared memory.
The \texttt{while} loop in Peterson's algorithm, in case its condition
does not hold, is another example of a process being blocked.

@@

\texttt{unlock} checks to see if any processes are waiting to get the lock.
If so, it uses the \texttt{head} and \texttt{tail}
methods from the \texttt{list} module to resume the first process that got
suspend and to remove its context from the list.
Selecting the first process is a design choice.  Another implementation could
have picked the last one, and yet another implementation could have used
\texttt{choose} to pick an arbitrary one.  Selecting the first is a common
choice in lock implementations as it prevents \emph{starvation}:
\index{starvation}
every process
gets a chance to obtain the lock (assuming every process eventually releases
it).  Chapter~\ref{ch:starvation} will talk more about starvation.

@@

\chapter{Reader/Writer Locks using Busy Waiting}
\index{reader/writer lock}

@@

While correct, it is not considered a good solution.
The solution is an example of what is called \emph{busy-waiting}
\index{busy waiting}
(aka spin-waiting):
\index{spin waiting}
processes spin in a loop until some desirable application-level condition is met.
The astute reader might wonder if obtaining the mutex itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the mutex is available
(see Figure~\ref{fig:spinlocks}).

@@

A good example of a resource with multiple units is the
so-called \emph{bounded buffer} (aka \emph{ring buffer}).
\index{ring buffer}
It is essentially
a queue implemented using a circular buffer
\index{circular buffer}
of a certain length and two pointers:
one where items are inserted and one from where items are extracted.  If the
buffer is full, processes that want to add more items have to wait.  In the usual
formulation, processes that want to extract an item when the buffer is empty
also have to wait.
This problem is known as the ``Producer/Consumer Problem.''

@@

Note the quite different usage of the \texttt{l\_} semaphores and \texttt{n\_}
semaphores.
People often call semaphores that are used as locks \emph{binary semaphores},
\index{binary semaphore}
as their counters only take on the values 0 and 1.
They protect critical sections.
The second type of semaphores are called \emph{counting semaphores}.
\index{counting semaphore}
They can be used to send signals between processes.
However, binary and counting semaphores are implemented the same way.

@@

In this chapter, we will present a version of a reader/writer lock
implementation that solves this type of starvation, but won't eliminate
the type of starvation.
In particular, the probability that either a reader or writer gets starved
will be exponentially diminishing.
We shall call such a solution \emph{fair}
\index{fairness}
(although it does not quite
match the usual formal nor vernacular concepts of fairness).

@@

The actor synchronization model is popular in a variety of programming
languages, including Erlang and Scala.  Actor support is also available
through popular libraries such as Akka, which is available for various
programming languages.  In Python, Java, and C/C++,
actors can be easily emulated using threads and \emph{synchronized queues}
\index{synchronized queue}
(aka \emph{blocking queues})
\index{blocking queue}
for messaging.
Each thread would have one such queue for receiving messages.
Dequeuing from an empty synchronized queue blocks the thread until
another thread enqueues a message on the queue.

@@

The solution is not practical because the \texttt{items} array must be
unbounded if an unbounded number of items may be produced.  Each slot
in the array is only used once, which is inefficient.  Also, the \texttt{inc}
and \texttt{swap} interlock instructions are not universally available
on existing hardware.
However, in the literature there are many examples of practical
non-blocking (aka \emph{wait-free})
\index{wait-free synchronization}
synchronization algorithms~\cite{}.

@@

An \emph{lvalue}
\index{lvalue}
(short for left hand value of an assignment statement)
is something that can be assigned.  This can be a
shared variable, a process variable, or a dereferenced pointer
variable.  It can also be indexed.  Examples include
\texttt{var}, \texttt{var[e]}, and \texttt{\^{}p}.

@@

A context is considered \emph{blocked}
\index{blocked process}
if there does not exist a
sequence instructions that the process can execute that either writes
to a shared variable (even if that leaves the shared variable unchanged)
or terminates the process.
